General constraints on magic distillation protocols from statistical mechanics

### 2021-05-11 magic_paper.tex

Then the magic distillation rate $R = m/n$ for the protocol is bounded by the expression
\begin{equation}\label{eq:rate_bounds_proof}
	R \leq \dfrac{\ln{\big( 1-\frac{4}{3}\epsilon \big)} + \beta (\phi - F)}{\ln{\big( 1-\frac{4}{3}\epsilon' \big)} + \beta (\phi' - F')},
\end{equation}
where $F$ is the free energy of $\tau$,  and 
\begin{equation}
	\phi = -\beta^{-1} \log \zeta
\end{equation}
with $\zeta$ given by the expressions
\begin{align}
	\zeta &= \sum_{k\in \mathbb{Z}_3} \alpha_k e^{-\beta E_k}, \nonumber\\
	\alpha_k &= \sum_{r \in \mathbb{Z}_3} \<E_k| A_{\x_\star} |E_k\>,
\end{align}
with $\x_\star = \mbox{argmin}[W_\tau( \x )]$. The primed variables are defined similarly for the output system.
\end{theorem}
The proof of this is provided in~\cref{free-energy-bound-proof}, and is a generalization of the analysis for unital protocols.

The above bound depends on: (a) quantum computational measures $\epsilon, \epsilon'$, (b) thermodynamic quantities $F, F',$  and (c) intermediate terms $\phi, \phi'$. The intermediate terms depend on how the energy eigenbasis of the system $\{|E_k\>\}$ relates to the computational stabilizer basis $\{|k\>\}$.  In particular, the energetic term $\phi$ quantifies the degree to which the negativity in the Wigner representation can be ascribed a sharp energetic value for the given Hamiltonian of the system. Its form is similar to $F$, and so can informally be viewed as a `magic free energy' term, associated to a `magic partition function' $\zeta$. While the coefficients $\alpha_k$ may be negative for some $k$, the quantity $\phi$ is always well-defined, since the function $\zeta$ is a positive multiple of the stabilizer Wigner component $W_\tau(0,0)$.

The sole purpose of the Clifford unitaries $C_1,C_2$ is to simplify the form of the $\alpha_k$ coefficients. For example, if $H=Z$ then the energy eigenbasis is equivalent to the computational basis up to a Fourier transform, and so $\phi$ would have a sharp, temperature-independent value $\phi \in \{E_0,E_1,E_2\}$ once we account for the Clifford change in basis.

@@

\begin{proposition}\label{prop:rmajor}
Given probability distributions $\bmx, \bmy, \r , \r'$, such that the components of $\r$ and $\r'$ are positive, the following statements are equivalent:
  \begin{enumerate}
    \item[(i)] $\bmy = A\bmx$ and $\r' = A\r$ for a stochastic map $A$;
    \item[(ii)]\label{en:tm3} $\sum\limits_{i=1}^n \abs{x_i - r_i t} \leq \sum\limits_{i=1}^n \abs{y_i - r'_i t}$ for all $t \in \mathbb{R}$;
    \item[(iii)] $L_{\bmx|\r}(t) \leq L_{\bmy|\r'}(t)$ for $t\in [0,1)$ and \vspace{5pt}\\ $L_{\bmx|\r}(1) = L_{\bmy|\r'}(1)$.
  \end{enumerate}
\end{proposition}
The proofs for these can be found in~\cite{cit:marshall,cit:bhatia,cit:nielsen,cit:lostaglio} and references therein.

The following result is used in the text to relate relative majorization of quasi-distributions to their Lorenz curves.
\begin{proposition}\label{lemma:Lorenz_linearity}
	Let $\p$ be a quasi-probability distribution and let $\r$ be a probability distribution with strictly non-zero components. Let $a > 0$ and $b \in \mathbb{R}$ then $L_{a\p + b \r | \r} (x) = a L_{\p |\r}(x) + b x$.
\end{proposition}
\begin{proof} 
	The Lorenz curve of $a\p + b \r$ relative to $\r$ passes through $(0,0)$ and the points $(\sum_{i=1}^k{r_{\pi(i)}}, \sum_{i=1}^k(a \p + b \r)_{\pi(i)})$ where $\pi$ is the permutation that puts $(a p_i/r_i + b)$ in non-increasing order. Since $a > 0$, the permutation $\pi$ puts  $(p_i/r_i)$ in non-increasing order too. We thus have
\begin{align*}
&\left( \sum_{i=1}^kr_{\pi(i)}, \sum_{i=1}^k(a \p + b \r)_{\pi(i)} \right) = \\ 
&\left( \sum_{i=1}^k r_{\pi(i)},a \sum_{i=1}^k  p_{\pi(i)} + b\sum_{i=1}^k r_{\pi(i)} \right) \nonumber,
\end{align*}
therefore the value of the Lorenz function at each potential elbow point $x_k = \sum_{i=1} ^kr_{\pi(i)}$ is given by
\begin{align}
&L_{a \p +b \r|\r} (x_k) = a L_{\p|\r} (x_k) + b L_{\r|\r}(x_k) = \nonumber\\
&a L_{\p|\r} (x_k) + b x_k,
\end{align}
so we have $L_{a\p  + b\r|\r} (x) = a L_{\p |\r}(x) + b x$ for any $x \in [0,1]$ due to linearity.
\end{proof}

Within the context of the Wigner representation, we define for convenience 
\begin{equation}
W_{\rho| \tau} (\x) := \frac{W_\rho (\x) }{ W_\tau(\x)},
\end{equation}
which is well-defined if $\tau$ is a full-rank stabilizer state.
\begin{theorem*}
	Given a magic state $\rho$, the maximum $L_\star$ of its Lorenz curve $\lc{\rho}{\sigma}(x)$ is independent of the $\sigma$--fragment and equal to $1+\sn{\rho}$. Moreover, the majorization constraint is stronger than mana in every fragment.
\end{theorem*}
\begin{proof}
	We denote the Wigner distributions of the states single-component vectors $\w(\rho)=W_\rho(x)$ and $ \w(\sigma)=W_\sigma(x)$. Likewise, we define
	\begin{equation}
\w(\rho|\sigma) = \frac{W_\rho (\x)}{W_{\sigma}(\x)}.
\end{equation}
	We choose the component indexing so that $\w(\rho|\sigma)^\downarrow = \w(\rho|\sigma)=:\tilde{\w}$ and so the components are sorted in non-increasing order.

@@

The following lemma shows that the re-scaled Lorenz curves behave in a natural way under tensor products. 
\begin{lemma}Let $\tau_A, \tau_B$ be full rank stabilizer states on systems $A$ and $B$, and let $\rho_A, \rho_B$ be arbitrary states on $A,B$. Then the rescaled quasi-distribution obeys
\begin{equation}
W_{\rho_A \otimes \rho_B | \tau_A \otimes \tau_B}(\bmx_A \oplus \bmx_B) =W_{\rho_A | \tau_A}(\x_A) W_{\rho_B | \tau_B}(\x_B)
\end{equation}
\end{lemma}
\begin{proof} This follows from the definitions. 
\begin{align}
W_{\rho_A \otimes \rho_B | \tau_A \otimes \tau_B}(\bmx_A \oplus \bmx_B) &= \frac{W_{\rho_A \otimes \rho_B} (\x_A \oplus \x_B)}{W_{\tau_A \otimes \tau_B} (\x_A \oplus \x_B)}\nonumber \\
 &= \frac{W_{\rho_A } (\x_A )W_{\rho_B } (\x_B )}{W_{\tau_A } (\x_A )W_{\tau_B } (\x_B )}\nonumber \\
  &= W_{\rho_A | \tau_A}(\x_A) W_{\rho_B | \tau_B}(\x_B)\nonumber,
\end{align}
as claimed.
\end{proof}

>>>>

	Let $T =(k\beta)^{-1}$ be any finite temperature for the physical system and let $H= \sum_{k \in \mathbb{Z}_3} E_k |E_k\>\<E_k|$ be the Hamiltonian of each qutrit subsystem in its eigen-decomposition.
Assume that in the thermodynamic limit ($n,m \gg 1$), the protocol applied to the equilibrium state $\tau^{\otimes n} = (e^{-\beta H}/\Z)^{\otimes n}$ maps $\tau^{\otimes n} \longrightarrow \tau^{\prime \otimes m}$, where we express state $\tau'$ as $\tau' = e^{-\beta H'}/\Z'$ for some Hermitian $H'$.

Then the magic distillation rate $R = m/n$ for the protocol is bounded by the expression
\begin{equation}\label{eq:rate_bounds_proof}
	R \leq \dfrac{\ln{\big( 1-\frac{4}{3}\epsilon \big)} + \beta (\phi - F)}{\ln{\big( 1-\frac{4}{3}\epsilon' \big)} + \beta (\phi' - F')},
\end{equation}
where $F$ is the free energy of $\tau$,  and 
\begin{equation}
	\phi = -\beta^{-1} \log \zeta
\end{equation}
with $\zeta$ given by the expressions
\begin{align}
	\zeta &= \sum_{k\in \mathbb{Z}_3} \alpha_k e^{-\beta E_k}, \nonumber\\
	\alpha_k &= \<E_k| A_{\x_\star} |E_k\>,
\end{align}
and $\x_\star = \arg\!\min W_\tau(\x)$. The primed variables are defined similarly for the output system.
\end{theorem}
\noindent The proof of this is provided in~\cref{free-energy-bound-proof}, and is a generalization of the analysis for unital protocols.

The above bound depends on: (a) quantum computational measures $\epsilon, \epsilon'$, (b) thermodynamic quantities $F, F',$  and (c) intermediate terms $\phi, \phi'$. The intermediate terms depend on how the energy eigenbasis of the system $\{|E_k\>\}$ relates to the computational stabilizer basis $\{|k\>\}$.  In particular, the energetic term $\phi$ quantifies the degree to which the negativity in the Wigner representation can be ascribed a sharp energetic value for the given Hamiltonian of the system. Its form is similar to $F$, and so can informally be viewed as a `magic free energy' term, associated to a `magic partition function' $\zeta$. The coefficients $\alpha_k$ may be negative for some $k$, when the Hamiltonian eigen-decomposes into non-stabilizer states, but the quantity $\phi$ is always well-defined, since the function $\zeta$ is a positive multiple of the stabilizer Wigner component $W_\tau(0,0)$, as seen in the proof of the theorem.

The sole purpose of the Clifford unitaries $C_1,C_2$ is to simplify the form of the $\alpha_k$ coefficients. \nick{That's not right, without Clifford processing we would get bounds that depend on $u$ whenever $\x_\star$ corresponds to $u$, as well as $v$ whenever $\x_\star$ corresponds to $v$ as seen in the contour plot - these are different values} For example, if $H=Z$ then the energy eigenbasis is equivalent to the computational basis up to a Fourier transform, and so $\phi$ would have a sharp, temperature-independent value $\phi \in \{E_0,E_1,E_2\}$ once we account for the Clifford change in basis.

@@

\begin{proposition}\label{prop:rmajor}
Given probability distributions $\bmx, \bmy$, $\r , \r'$, such that the components of $\r$ and $\r'$ are positive, the following statements are equivalent:
  \begin{enumerate}
    \item[(i)] $\bmy = A\bmx$ and $\r' = A\r$ for a stochastic map $A$;
    \item[(ii)]\label{en:tm3} $\sum\limits_{i=1}^n \abs{x_i - r_i t} \leq \sum\limits_{i=1}^n \abs{y_i - r'_i t}$ for all $t \in \mathbb{R}$;
    \item[(iii)] $L_{\bmx|\r}(t) \leq L_{\bmy|\r'}(t)$ for $t\in [0,1)$ and \vspace{5pt}\\ $L_{\bmx|\r}(1) = L_{\bmy|\r'}(1)$.
  \end{enumerate}
\end{proposition}
The proofs for these can be found in~\cite{cit:marshall,cit:bhatia,cit:nielsen,cit:lostaglio} and references therein.

The following result is used in the text to relate relative majorization of quasi-distributions to their Lorenz curves.
\begin{proposition}\label{lemma:Lorenz_linearity}
	Let $\p$ be a quasi-probability distribution and let $\r$ be a probability distribution with strictly non-zero components. Let $a > 0$ and $b \in \mathbb{R}$ then $L_{a\p + b \r | \r} (x) = a L_{\p |\r}(x) + b x$.
\end{proposition}
\begin{proof} 
	The Lorenz curve of $a\p + b \r$ relative to $\r$ passes through $(0,0)$ and the points $(\sum_{i=1}^k{r_{\pi(i)}}, \sum_{i=1}^k(a \p + b \r)_{\pi(i)})$ where $\pi$ is the permutation that puts $(a p_i/r_i + b)$ in non-increasing order. Since $a > 0$, the permutation $\pi$ puts  $(p_i/r_i)$ in non-increasing order too. We thus have
\begin{align*}
&\left( \sum_{i=1}^kr_{\pi(i)}, \sum_{i=1}^k(a \p + b \r)_{\pi(i)} \right) = \\ 
&\left( \sum_{i=1}^k r_{\pi(i)},a \sum_{i=1}^k  p_{\pi(i)} + b\sum_{i=1}^k r_{\pi(i)} \right) \nonumber,
\end{align*}
so the value of the Lorenz curve at each potential elbow point $x_k = \sum_{i=1} ^kr_{\pi(i)}$ is given by
\begin{align}
&L_{a \p +b \r|\r} (x_k) = a L_{\p|\r} (x_k) + b L_{\r|\r}(x_k) = \nonumber\\
&a L_{\p|\r} (x_k) + b x_k,
\end{align}
so we have $L_{a\p  + b\r|\r} (x) = a L_{\p |\r}(x) + b x$ for any $x \in [0,1]$ due to linearity.
\end{proof}

Within the context of the Wigner representation, we define the re-scaled quasi-distribution
\begin{equation}
	W_{\rho| \tau} (\x) := \frac{W_\rho (\x) }{ W_\tau(\x)},
\end{equation}
which is well-defined if $\tau$ is a full-rank stabilizer state.

The following lemma shows that the re-scaled Lorenz curves behave in a natural way under tensor products. 
\begin{proposition}
	Let $\tau_A, \tau_B$ be full rank stabilizer states on systems $A$ and $B$, and let $\rho_A, \rho_B$ be arbitrary states on $A,B$. Then the rescaled quasi-distribution obeys
\begin{equation}
W_{\rho_A \otimes \rho_B | \tau_A \otimes \tau_B}(\bmx_A \oplus \bmx_B) =W_{\rho_A | \tau_A}(\x_A) W_{\rho_B | \tau_B}(\x_B)
\end{equation}
\end{proposition}
\begin{proof} This follows from the definitions. 
\begin{align}
W_{\rho_A \otimes \rho_B | \tau_A \otimes \tau_B}(\bmx_A \oplus \bmx_B) &= \frac{W_{\rho_A \otimes \rho_B} (\x_A \oplus \x_B)}{W_{\tau_A \otimes \tau_B} (\x_A \oplus \x_B)}\nonumber \\
 &= \frac{W_{\rho_A } (\x_A )W_{\rho_B } (\x_B )}{W_{\tau_A } (\x_A )W_{\tau_B } (\x_B )}\nonumber \\
  &= W_{\rho_A | \tau_A}(\x_A) W_{\rho_B | \tau_B}(\x_B)\nonumber,
\end{align}
as claimed.
\end{proof}

\begin{theorem*}
	Given a magic state $\rho$, the maximum $L_\star$ of its Lorenz curve $\lc{\rho}{\sigma}(x)$ is independent of the $\sigma$--fragment and equal to $1+\sn{\rho}$. Moreover, the majorization constraint is stronger than mana in every fragment.
\end{theorem*}
\begin{proof}
	We denote the Wigner distributions of the states single-component vectors $\w(\rho)=W_\rho(x)$ and $ \w(\sigma)=W_\sigma(x)$. Likewise, we define
	\begin{equation}
\w(\rho|\sigma) = \frac{W_\rho (\x)}{W_{\sigma}(\x)}.
\end{equation}
	We choose the component indexing so that $\w(\rho|\sigma)^\downarrow = \w(\rho|\sigma)=:\tilde{\w}$ and so the components are sorted in non-increasing order.

@@
